Suitability of Graph Neural Network for Shape Memory Alloys

### 2024-05-22 src/tex/ms.tex
First Major Refactor of Draft Preprint (#1)
Not the build failed by intention as I cannot run all the simulations on Github actions.

## Refactors
* Rename the EAM potential files.

* Refactor minimize, EOS, elastic, etc.

* Refactor function names. Core should work!

* Refactored Gruneisen code and coh. energy plot

* MEAM by Ko et al. potential files setup, not validated but working.

* Calculation scripts good. Some plotting good. Added draft MD simulation script. New Calculator.

* Fairly complete refactor. Working Process.

* New scripts, potential, rework Snakemake, move folders, added m

@@

% Abstract with filler text
\begin{abstract}
	Shape Memory Alloys (SMAs) such as NiTi are pivotal in vibrational
	dampening,robotic arms, prothestic hands, and mechanical vales as a
	result of their unique displacive phase transformation
	properties. However, assessing their phonon characteristics, which
	provide insight into the displacive phase transformation, by using
	conventional Density Functional Theory (DFT) methods is
	computationally challenging. In this paper we assess the use of
	recent Graph Neural Network (GNN) potentials as a viable
	alternative. These GNN potentials have leveraged large DFT-computed
	databases for training and provide a good approximation to the
	potential energy surfaces which enable inference on energies and
	forces in lattice and atomic dynamic calculations. We compare GNN
	potentials with the Embedded Atom Method (EAM) and Modified EAM
	style potentials in predicting phonon properties of NiTi and
	PtTi. We find that in general the EAM/MEAM potentials are slightly
	more performant than the GNN in that they appear to more accurately
	describe equation of state, phonon dispersions, and elastic
	constants. However, the MACE potential shows significant promise as
	an exploritory potential for designing and characterizing more
	complex SMAs. Our findings suggest GNN potentials as a promising
	initial avenue for advancing SMA research, in particular ternary and
	quaternary systems, upon fine-tuning of pre-trained GNN models with
	more relevant DFT data (i.e., force-constants) to permit better
	outcomes.
\end{abstract}

@@

Recently, there as been a growing effort in the development of machine
learning potentials. These deviate from traditional physics approaches
to interatomic potentials, which have historically been inspired from
more accurate methods like tight-binding method. In machine learning
potentials, one describes the atoms and structure using descriptors of
features, these are typically learned or prescribed. The features can
be prescribed in isolation or combination of structural,
compositional, and property characteristics of the material data
point. Futhermore, one may choose to fined new representations via
unsurpervised learning. \par

The subsequent step once the featureization process is completed is to
determine a suitable neural network (NN) architecture which can learn
the target property or properties given the feature inputs. In the
case of atomistic simulations the target is prediction of the
potential energy surface and corresponding gradients (i.e.,
forces). Two of the most early attempts to do this were the works of
\citet{Bartok2010} and \citet{Thompson2015}, whereby projection onto
higher-dimensional basis, such as bispectrum components of local
neighbor density, is learned in order to describe the atomic
environments. The energy per atom is then determined by these learned
features by employing linear regression or non-linear Gaussian process
regression. We note that these are just two of the early examples but
similar potentials have been developed in earlier, in parallel, or
recently by various groups
\cite{Behler2007,Wang2018,Deringer2019,Lee2019}. \par

While these early attempts using ML models were successfully, they may
only include limited representations of structural and/or chemical
details to learn new features from which could limit
transferability. Futhermore, the training process was typically
restricted to a limited number of elemental constieunts of the target
material system. As a consequence, for each chemical system and phases
of interest one needs to provide a suitable training dataset. To move
beyond such limitations, groups began using graph neural network (GNN)
architectures and message passing scheme to represent the atomic
structures as nodes and edges \cite{Reiser2022}. This enabled two
distinctions, first being that the nodes and edges corresponding to
the atoms and bonds could include features, and secondly the learning
could be done over a large structural and compositional space enabling
a generalization\footnote{Typically these are referred to as Universal
	interatomic potentials or forcefields since they in theory
	generalize well to any chemical system or phase.}. The use of GNNs
is efficient in representing structure-composition-property inputs to
additional NN layers for predicting potential energy surfaces and
gradients and was initially pursued in the CGCNN, MEGNet, and ALIGNN
potentials \cite{Choudhary2021,Chen2019,Xie2018}. The differences
between the approaches are predominately in atomic and bond features,
nearest-neighbor graph, and specific NN archetcture. \par

The training of these early GNN utlized large density functional
theory calculation repositories and mostly tuned the learning cost
function to ground-state structure and energy. This resulted in
mean-absolute-errors and other measures of performance that could be
improved upon \cite{Riebesell2023}. More recently there have been
significant improvements in both training data used and the GNN
architecture that show promise for truly approaching a general
interatomic potential suitable for static and dynamic atomistic
calculations of multicomponents and multiphases. If succesfull,
identifying and designing new potential SMA materials using GNN
interatomic potentials could have tremendous impact on
technology. Therefore, the primary goal is to provide a qualitative
assessment of pre-trained GNN interatomic potentials that are promoted
as being transferable across compositions and phases. We specfically
focus on the predictions for equations of state, phonon bandstructure,
and elastic constants for various phases of NiTi and PtTi using three
GNN potentials: M3GNet, CHGNet, and MACE
\cite{Chen2022,Deng2023,Batatia2022}. We note that the
\citet{Riebesell2023} provides a summarized performance over these GNN
using the materials project database, but does not specifically
address SMA and phonon characteristics. \par

@@

\subsection{Workflow}
\label{sec:workflow}
To promote reproducible research and align with findable, accessible,
interoper-able and reusable (FAIR) principles\cite{Walsh2024}, we
utilize \showyourwork python workflow\cite{Luger2021} and several open
source python packages. Instructions for reproducing calculated data,
figures, tables, and manuscript are provided in
appendix~\ref{sec:appx_reproduce} and corresponding github
repo\footnote{\url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP}}. We
note that in order to achieve the different computing environments
required by the interatomic potentials we leverage the ability to
specify the python computing environments using
\href{https://github.com/showyourwork/showyourwork}{\showyourwork}
package.\par

\subsection{Calculation Details}
\label{sec:calc_details}
The ground-state structures for the NiTi phases \textit{B2},
\textit{B19}, \textit{B19'}, and \textit{BCO} are generated using the
atomic simulation environment (ASE) package \cite{Larsen2017}. This
requires that the unit cell parameters, basis atoms, and spacegroup
number be provided. The initial unit cell parameters and basis
corrdinates for these structures have been taken from the materials
project \cite{Jain2013} and several references
\cite{Haskins2016,Kadkhodaei2018}. The initial structures for NiTi are
shown in fig~\ref{fig:niti_structures} and are adapted for PtTi as
well given same space group symmetries. The energy and forces on the
atoms are determined using the ASE calculator interfaces for EAM,
MEAM, M3GNet, CHGNet, and
MACE.\cite{Mutter2010,Zhong2011,Ko2015,Kim2017,Chen2022,Deng2023,Batatia2022}
The EAM and MEAM potential interfaces are achieved through the use of
LAMMPS python library \cite{Thompson2022} and ASE. We have also
implemented calculators for potentials by \cite{Kavousi2019},
\cite{Wang2018}, and \cite{Choudhary2021} but are not shown in the
main manuscript of this work.\par

@@

\begin{figure}[ht!]
	\label{fig:opt_density}
	\centering
	\includegraphics[width=\linewidth]{figures/NiTi_Equilibrium_Density.png}
	\caption{Equilibrium Density for \ce{NiTi} structures with different
		models.}
\end{figure}

@@

With $E_{\text{bulk}}$ the total energy of the system,
$\sum_i^N n_i E_{\text{atom}_i}$ is the sum over isolated atom
energies for $n_i$ species, and $N$ the number of atoms in bulk
reference. Classical interatomic potentials are offset with respect to
the isolated atom energies, so these terms are zero and neglected,
giving $E_{c} = \frac{E_{\text{bulk}}}{N}$. In fig.~\ref{fig:ecoh} the
bar chart for different phases and interatomic potentials models are
show. We compare these to dervided experimental values for
\textit{B2}\cite{Vandermause2024} and the value from The Materials
Project (MP) entry \texttt{MP-571} which corresponds to the
\textit{B2} phase \cite{MP--571}. In the figure we see that the EAM
and MEAM potentials show significantly lower cohesive energy compared
to the GNN potentials and are in better agreement with the
experimental value. Previous studies using the EAM and MEAM potentials
have been shown to reproduce previous experimental and DFT data
\cite{Haskins2016,Ko2015} such as cohesive energy. The GNNs predict a
more strongly bound and stable \textit{B2} structure but all are
consistent with the entry for The MP and calculation details can be
found in ref.~\cite{MP--571}. This consistency is resonable
considering all pretrained GNN models used in this work have been
developed with MP data. \par

@@

\begin{figure}[!htp]
	\script{GruneisenParameters.py} \centering
	\includegraphics[width=\columnwidth]{figures/Plot_NiTi_M_ModeGruneisen.png}
	\label{fig:modegruneisen}
	\caption{The sign-log of the \textit{M}-mode Gr\"{u}neisen
	parameter, $\gamma$, as a function of eigenfrequency for different
	interatomic potentials.}
\end{figure}

@@

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_M3GNet_B2_StrainsPhononBandstructures.png} \vspace{1mm}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_CHGNet_B2_StrainsPhononBandstructures.png} \vspace{1mm}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_MACE_B2_StrainsPhononBandstructures.png}
		\caption{ Phonon dispersion of B2 structure for the (top) M3GNet,
			(middle) CHGNet, and (bottom) MACE potentials for different
			isotropic strains. Similar transcparency is used as done in
			\ref{fig:mutter_zhong_phonon_b2}.  }
		\label{fig:gnn_phonon_b2}
	\end{centering}
\end{figure}

@@

TODO

>>>>

\usepackage{pdflscape}

  % Customize footer to show only "preprint"
\makeatletter
\def\ps@pprintTitle{%
     \let\@oddhead\@empty
     \let\@evenhead\@empty
     \def\@oddfoot{\hfill\footnotesize\itshape Preprint, Incomplete Draft\hfill}%
     \let\@evenfoot\@oddfoot}
\makeatother

@@

% Abstract with filler text
\begin{abstract}
	Shape Memory Alloys (SMAs) such as NiTi are pivotal in vibrational
	dampening,robotic arms, prosthetic hands, and mechanical vales as a
	result of their unique displacive phase transformation
	properties. However, assessing their phonon characteristics, which
	provide insight into the displacive phase transformation, by using
	conventional Density Functional Theory (DFT) methods is
	computationally challenging. In this paper we assess the use of
	recent Graph Neural Network (GNN) potentials as a viable
	alternative. These GNN potentials have leveraged large DFT-computed
	databases for training and provide a good approximation to the
	potential energy surfaces which enable inference on energies and
	forces in lattice and atomic dynamic calculations. We compare GNN
	potentials with the Embedded Atom Method (EAM) and Modified EAM
	style potentials in predicting phonon properties of NiTi and
	PtTi. We find that in general the EAM/MEAM potentials are slightly
	more performant than the GNN in that they appear to more accurately
	describe equation of state, phonon dispersions, and elastic
	constants. However, the MACE potential shows significant promise as
	an exploratory potential for designing and characterizing more
	complex SMAs. Our findings suggest GNN potentials as a promising
	initial avenue for advancing SMA research, in particular ternary and
	quaternary systems, upon fine-tuning of pre-trained GNN models with
	more relevant DFT data (i.e., force-constants) to permit better
	outcomes.
\end{abstract}

@@

Recently, there as been a growing effort in the development of machine
learning potentials. These deviate from traditional physics approaches
to interatomic potentials, which have historically been inspired from
more accurate methods like tight-binding method. In machine learning
potentials, one describes the atoms and structure using descriptors of
features, these are typically learned or prescribed. The features can
be prescribed in isolation or combination of structural,
compositional, and property characteristics of the material data
point. Futhermore, one may choose to fined new representations via
unsupervised learning. \par

The subsequent step once the featurization process is completed is to
determine a suitable neural network (NN) architecture which can learn
the target property or properties given the feature inputs. In the
case of atomistic simulations the target is prediction of the
potential energy surface and corresponding gradients (i.e.,
forces). Two of the most early attempts to do this were the works of
\citet{Bartok2010} and \citet{Thompson2015}, whereby projection onto
higher-dimensional basis, such as bispectrum components of local
neighbor density, is learned in order to describe the atomic
environments. The energy per atom is then determined by these learned
features by employing linear regression or non-linear Gaussian process
regression. We note that these are just two of the early examples but
similar potentials have been developed in earlier, in parallel, or
recently by various groups
\cite{Behler2007,Wang2018,Deringer2019,Lee2019}. \par

While these early attempts using ML models were successfully, they may
only include limited representations of structural and/or chemical
details to learn new features from which could limit
transferability. Futhermore, the training process was typically
restricted to a limited number of elemental constituents of the target
material system. As a consequence, for each chemical system and phases
of interest one needs to provide a suitable training dataset. To move
beyond such limitations, groups began using graph neural network (GNN)
architectures and message passing scheme to represent the atomic
structures as nodes and edges \cite{Reiser2022}. This enabled two
distinctions, first being that the nodes and edges corresponding to
the atoms and bonds could include features, and secondly the learning
could be done over a large structural and compositional space enabling
a generalization\footnote{Typically these are referred to as Universal
	interatomic potentials or forcefields since they in theory
	generalize well to any chemical system or phase.}. The use of GNNs
is efficient in representing structure-composition-property inputs to
additional NN layers for predicting potential energy surfaces and
gradients and was initially pursued in the CGCNN, MEGNet, and ALIGNN
potentials \cite{Choudhary2021,Chen2019,Xie2018}. The differences
between the approaches are predominately in atomic and bond features,
nearest-neighbor graph, and specific NN architecture. \par

The training of these early GNN utilized large density functional
theory calculation repositories and mostly tuned the learning cost
function to ground-state structure and energy. This resulted in
mean-absolute-errors and other measures of performance that could be
improved upon \cite{Riebesell2023}. More recently there have been
significant improvements in both training data used and the GNN
architecture that show promise for truly approaching a general
interatomic potential suitable for static and dynamic atomistic
calculations of multi-components and multi-phases. If successful,
identifying and designing new potential SMA materials using GNN
interatomic potentials could have tremendous impact on
technology. Therefore, the primary goal is to provide a qualitative
assessment of pre-trained GNN interatomic potentials that are promoted
as being transferable across compositions and phases. We specifically
focus on the predictions for equations of state, phonon bandstructure,
and elastic constants for various phases of NiTi and PtTi using three
GNN potentials: M3GNet, CHGNet, and MACE
\cite{Chen2022,Deng2023,Batatia2022}. We note that the
\citet{Riebesell2023} provides a summarized performance over these GNN
using the materials project database, but does not specifically
address SMA and phonon characteristics. \par

@@

\subsection{Workflow}
\label{sec:workflow}
To promote reproducible research and align with findable, accessible,
interoperable and reusable (FAIR) principles\cite{Walsh2024}, we
utilize \showyourwork python workflow\cite{Luger2021} and several open
source python packages. Instructions for reproducing calculated data,
figures, tables, and manuscript are provided in
appendix~\ref{sec:appx_reproduce} and corresponding github
repo\footnote{\url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP}}. We
note that in order to achieve the different computing environments
required by the interatomic potentials we leverage the ability to
specify the python computing environments using
\href{https://github.com/showyourwork/showyourwork}{\showyourwork}
package.\par

\subsection{Calculation Details}
\label{sec:calc_details}
The ground-state structures for the NiTi phases \textit{B2},
\textit{B19}, \textit{B19'}, and \textit{BCO} are generated using the
atomic simulation environment (ASE) package \cite{Larsen2017}. This
requires that the unit cell parameters, basis atoms, and spacegroup
number be provided. The initial unit cell parameters and basis
coordinates for these structures have been taken from the materials
project \cite{Jain2013} and several references
\cite{Haskins2016,Kadkhodaei2018}. The initial structures for NiTi are
shown in fig~\ref{fig:niti_structures} and are adapted for PtTi as
well given same space group symmetries. The energy and forces on the
atoms are determined using the ASE calculator interfaces for EAM,
MEAM, M3GNet, CHGNet, and
MACE.\cite{Mutter2010,Zhong2011,Ko2015,Kim2017,Chen2022,Deng2023,Batatia2022}
The EAM and MEAM potential interfaces are achieved through the use of
LAMMPS python library \cite{Thompson2022} and ASE. We have also
implemented calculators for potentials by \cite{Kavousi2019},
\cite{Wang2018}, and \cite{Choudhary2021} but are not shown in the
main manuscript of this work.\par

@@

\begin{figure}[ht!]
	\script{PlotDensity.py}
	\centering
	\includegraphics[width=\linewidth]{figures/NiTi_Equilibrium_Density.png}
	\caption{Equilibrium Density for \ce{NiTi} structures with different
		models.}
	\label{fig:opt_density}
\end{figure}

@@

With $E_{\text{bulk}}$ the total energy of the system,
$\sum_i^N n_i E_{\text{atom}_i}$ is the sum over isolated atom
energies for $n_i$ species, and $N$ the number of atoms in bulk
reference. Classical interatomic potentials are offset with respect to
the isolated atom energies, so these terms are zero and neglected,
giving $E_{c} = \frac{E_{\text{bulk}}}{N}$. In fig.~\ref{fig:ecoh_niti} the
bar chart for different phases and interatomic potentials models are
show. We compare these to dervided experimental values for
\textit{B2}\cite{Vandermause2024} and the value from The Materials
Project (MP) entry \texttt{MP-571} which corresponds to the
\textit{B2} phase \cite{MP--571}. In the figure we see that the EAM
and MEAM potentials show significantly lower cohesive energy compared
to the GNN potentials and are in better agreement with the
experimental value. Previous studies using the EAM and MEAM potentials
have been shown to reproduce previous experimental and DFT data
\cite{Haskins2016,Ko2015} such as cohesive energy. The GNNs predict a
more strongly bound and stable \textit{B2} structure but all are
consistent with the entry for The MP and calculation details can be
found in ref.~\cite{MP--571}. This consistency is reasonable
considering all pretrained GNN models used in this work have been
developed with MP data. \par

@@

\begin{figure}[!htp]
	\script{GruneisenParameters.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/Plot_NiTi_M_ModeGruneisen.png}
	\caption{The sign-log of the \textit{M}-mode Gr\"{u}neisen
	parameter, $\gamma$, as a function of eigenfrequency for different
	interatomic potentials.}
	\label{fig:modegruneisen}
\end{figure}

@@

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_M3GNet_B2_StrainsPhononBandstructures.png} \vspace{1mm}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_CHGNet_B2_StrainsPhononBandstructures.png} \vspace{1mm}
		\includegraphics[width=0.5\linewidth]{figures/NiTi_MACE_B2_StrainsPhononBandstructures.png}
		\caption{ Phonon dispersion of B2 structure for the (top) M3GNet,
			(middle) CHGNet, and (bottom) MACE potentials for different
			isotropic strains. Similar transcparency is used as done in
			\ref{fig:mutter_zhong_ko_phonon_b2}.  }
		\label{fig:gnn_phonon_b2}
	\end{centering}
\end{figure}

@@
